\vspace{-0.8in}Brief Announcement: A Family of Leaderless Generalized-Consensus Algorithms

### 2016-02-12 podc16/paper2.tex
work on the intro

We show that the core idea underlying EPaxos can be captured in a generic algorithm that uses two abstractions: (1), a per-command dependency-set computation algorithm, a concept that we define in the paper, and (2), an algorithm ensuring, for each command proposed to it, agreement on a dependency set. Both abstractions lend themselves well to implementations that avoid the use of a unique leader process.

Our abstract algorithm gives rise to a family of algorithms whose members are obtained by using a concrete map-agreement algorithm and a concrete dependency-set computation algorithm. 
Note that any generalized consensus algorithm can be used to implement (2), but only a weaker problem than generalized consensus needs to be solved because no ordering among proposal is required.

Our work has two main advantages: it enables modular correctness proofs of algorithms like EPaxos and it allows to tune the performance of implementations by changing the underlying map-agreement and dependency-set algorithms. We expect that these two tunable knobs will allow a principled theoretical and empirical evaluation of the trade-offs that can be achieved by this new family of algorithms.

@@

Our leaderless generalized consensus algorithm orchestrates a dependency-set algorithm and map-agreement algorithm as shown in \cref{fig:deps-set-algo3}.
A process $p$ maintains two local variables $m_p$ (a map from commands to sets of commands) and $\sigma_p$ (a c-struct). Upon  receiving a command $c$, $p$ proceeds in three phases. First, it calls a \textit{dependency-set} algorithm (\Cref{dep-algo}) to determine a set of commands $\mathcal{D}$ that the command $c$ may depend on. Second, $p$ proposes the mapping  $c\mapsto \mathcal{D}$ to a \textit{map-agreement} algorithm (\Cref{map-algo}); the map-agreement algorithm may subsequently commit a mapping $c\mapsto \mathcal{D}'$ for $c$, in which case $p$ inserts the mapping $\left[c \mapsto \mathcal{D}'\right]$ in $m_p$, and we say that $\mathcal{D}'$ is the dependency set of $c$.
We say that $c$ is executable in $m_p$ when all its transitive dependencies are known. %no command $c'$ for which no mapping has been committed is reachable from $c$ by recursively following committed mappings. 
Finally, when $c$ is executable in $m_p$, then $p$ may execute $c$ by deciding the c-struct obtained by \emph{locally} running a \emph{graph-processing} algorithm described in \Cref{abstract-algo}.
Any other process can run the three phases independently of $p$ and eventually execute the command despite a crash of $p$.
We show that the leaderless  generalized consensus algorithm guarantees the safety and liveness properties of generalized consensus when its underlying dependency-set and map-agreement algorithms satisfy their safety and liveness properties.

@@

A dependency-set algorithm exposes the input interface $\opn{announce}_p\left( c \right)$, to announce a command $c$, the output interface $\opn{suggest}_p \left( c,\mathcal{D}\right)$, to suggest a set of dependencies $\mathcal{D}$ for $c$, and an input interface $\opn{commit}\left( c,\mathcal{D} \right)$ (meant to observe the map-agreement algorithm), where $p\in P$ is the process executing the call.
Moreover, a dependency-set algorithm must ensure that:
\begin{itemize}[noitemsep,nolistsep]
    \item[-] Safety: (\setword{S1}{dep-valid}), for any suggestion $\opn{suggest}_p \left( c,\mathcal{D}\right)$, every command in $\{c\}\cup \mathcal{D}$ must have been announced before; (\setword{S2}{suggestion-agreement}), for any two suggestions $\opn{suggest}_p \left( c_1,\mathcal{D}_1\right)$ and $\opn{suggest}_q \left( c_2,\mathcal{D}_2\right)$, if $c_1$ and $c_2$ do not commute then either $c_1\in \mathcal{D}_2$ or $c_2 \in \mathcal{D}_1$.
%    \item If $c$ was announced by $p$, then eventually a dependency set is suggested for $c$ to $p$.
    \item[-] Liveness: (\setword{L1}{eventual-suggestion}) If a suggestion $\opn{suggest}_p\left( c \right)$ was made, then eventually a suggestion $\opn{suggest}\left( c,\mathcal{D} \right)_p$ is made, and (\setword{L2}{re-suggest}), if a suggestion $\opn{suggest}_p(c,\mathcal{D})$ has been made and no input $\opn{commit}_q\left( c,\mathcal{D'} \right)$ is observed, then eventually a new suggestion $\opn{suggest}_r\left( c,\mathcal{D''} \right)$ for $r\neq p$ is made.
\end{itemize}

An example of implementation of such an algorithm is depicted by the pseudocode of \cref{fig:deps-set-algo1,fig:deps-set-algo2}. Due to space constraints, our example focuses on satisfying the safety property only. The main idea is the following: a process $p$ announcing a command $c$ broadcasts $c$ to all the processes, and it suggests the dependency-set $deps_p[c]$ for $c$ as the union of the sets $\mathcal{D}$ that it receives as replies from a quorum of $\mathcal{Q}$ processes; on the other hand, a process that receives an announcement for $c$, computes its reply $\mathcal{D}$ for $c$ as the set of commands $d$ that it received so far, and such that $c$ and $d$ do not commute. The choice of the value of $\mathcal{Q}$ must be such that the intersection of any two quorums is not empty. About correctness, we can easily prove that such an implementation satisfies the safety property by noticing that:  any process $p$ inserts a command $d$ in a dependency-set $\mathcal{D}$ that $p$ sent as a reply for a command $c$, only if $p$ previously received $d$. Therefore, every command suggested in a dependency-set must have been announced before. Further, since a process can suggest a set of dependencies $deps_p[c]$ for a command $c$ after it has received a reply for $c$ from at least a quorum of processes, and thanks to the requirement that we enforced on the intersection of quorums, we can conclude that for any two suggestions $\opn{suggest}_p \left( c_1,deps_p[c_1]\right)$ and $\opn{suggest}_q \left( c_2,deps_q[c_2]\right)$,  if $c_1$ and $c_2$ do not commute then either $c_1\in deps_q[c_2]$ or $c_2 \in deps_p[c_1]$.

@@

		\begin{algorithmic}[1]
			{\scriptsize
			\START[$\opn{init}_p$]{{\bf upon}}{$()$}
			\State $m_p\gets \emptyset$
			\State $\sigma_p\gets \bot$
			\END
			\line(1,0){80}
			\START[$\opn{GC-propose}_p$]{{\bf upon}}{$(c)$}
			\State {\bf call} $\opn{announce}_p\left( c \right)$
			\END
			\line(1,0){80}
			\START[$\opn{suggest}_p$]{{\bf upon}}{$(c,\mathcal{D})$}
			\State  $\opn{propose}_p( c,\mathcal{D})$ 			
			\END
			\line(1,0){80}
			\START[$\opn{commit}_p$]{{\bf upon}}{$(c,\mathcal{D})$}
			\State  $m_p \gets m_p \oplus \left[ c \mapsto \mathcal{D}\right]$	
			\State $\sigma_p \gets  \opn{c-struct}\left(m_p\right)$
			\State $\opn{GC-decide}_p(\sigma_p)$		
			\END
			}
		\end{algorithmic}
	\end{varwidth}%
}
\label{fig:deps-set-algo3}
}
\hspace*{0.1in}\subfigure[Example of dependency-set algorithm implementation - part 1.]{
\noindent\fbox{%
	\begin{varwidth}{\dimexpr\linewidth-2\fboxsep-2\fboxrule\relax}

		\begin{algorithmic}[1]
			{\scriptsize
			\START[\texttt{init$_p$}]{{\bf upon}}{$()$} \label{alg:init}
			\State $cmds_p\gets\emptyset$ \label{alg:init-cmds}
			\State $\forall c,$ $deps_p[c]\gets \emptyset$ \label{alg:init-deps}
			\State $\forall c,$ $heard_p[c]\gets \emptyset$ \label{alg:init-heard}
			\State $\forall c,$ $suggested_p[c]\gets false$ \label{alg:init-suggested}
			\END
			\line(1,0){108}
			\START[\texttt{announce$_p$}]{{\bf upon}}{$(c)$} \label{alg:announce}
			\State $cmds_p\gets cmds_p\cup\{c\}$ \label{alg:add-command}
			\State $\forall q\in P$ {\bf send} $\langle \texttt{cmd}$ $|$ $c \rangle$ {\bf to} $q$\label{alg:send-new-cmd}
			\END
			}
		\end{algorithmic}
	\end{varwidth}%
}
\label{fig:deps-set-algo1}
}	
\hspace*{0.1in}\subfigure[Example of dependency-set algorithm implementation - part 2.]{
\noindent\fbox{%
	\begin{varwidth}{\dimexpr\linewidth-2\fboxsep-2\fboxrule\relax}

		\begin{algorithmic}[1]
			\makeatletter
			\setcounter{ALG@line}{8}
			\makeatother
			{\scriptsize
			\START[$\langle \texttt{cmd}$ $|$ $c \rangle$]{{\bf upon receive}}{ {\bf from} $q$} \label{alg:receive-new-cmd}
			\State $\mathcal{D} \gets \{ d : d \in cmds_p \land$\label{alg:update-d}
			\Statex \hspace{2cm}$\land d\neq c \land d \asymp c \}$
			\State $cmds_p\gets cmds_p\cup\{c\}$\label{alg:update-cmds-p}
			\State {\bf send} $\langle \texttt{deps}$ $|$ $c,\mathcal{D} \rangle$ {\bf to} $q$\label{alg:send-deps}
			\END
			\line(1,0){125}
			\START[$\langle \texttt{deps}$ $|$ $c,\mathcal{D} \rangle$]{{\bf upon receive}}{ {\bf from} $q$} \label{alg:receive-deps}
			\State $deps_p[c]\gets deps_p[c]\cup \mathcal{D}$\label{alg:update-deps-p}
			\State $heard_p[c]\gets heard_p[c]\cup \{q\}$
			\END
			\line(1,0){125}
			\START[]{{\bf upon}}{$\exists c$ $:$ $ | heard_p[c] | \geq \mathcal{Q}$ $\land$} \label{alg:receive-deps}
			\Statex \hspace{1cm}$\land$ $suggested_p[c]=false$
			\State $suggested_p[c]\gets true$
			\State {\bf call} $\texttt{suggest}_p(c,deps_p[c])$ \label{alg:call-suggest}
			\END
			}
		\end{algorithmic}
	\end{varwidth}%
}
\label{fig:deps-set-algo2}
}
\caption{Leaderless generalized consensus, and example of dependency-set algorithm implementation.}
\label{fig:deps-set-algo}
\end{figure*}

>>>>

We show that the core idea underlying EPaxos can be captured in a generic algorithm that uses two abstractions: a dependency-set algorithm, which suggests dependencies for commands, and a map-agreement algorithm, which ensures that, for each submitted command, processes agree on a dependency set. Both abstractions lend themselves well to implementations that avoid the use of a unique leader process.
Our generic algorithm gives rise to a family of algorithms whose members are obtained by using a concrete agreement algorithm and a concrete dependency-set algorithm. 
%Note that any generalized consensus algorithm can be used to implement (2), but only a weaker problem than generalized consensus needs to be solved because no ordering among proposal is required.

On top of enabling modular correctness proofs of algorithms like EPaxos, we expect that the modular structure of our generic leaderless algorithm will allow a principled theoretical and empirical evaluation of the trade-offs that can be achieved by EPaxos-like algorithms.

@@

Our leaderless generalized consensus algorithm orchestrates a dependency-set sub-algorithm and map-agreement sub-algorithm as shown in \cref{fig:deps-set-algo3}.
A process $p$ receiving a command $c$ proceeds in three phases. First, it calls a \textit{dependency-set} sub-algorithm (\Cref{dep-algo}) to determine a set of commands $\mathcal{D}$ that the command $c$ may depend on. Second, $p$ proposes the mapping  $c\mapsto \mathcal{D}$ to a \textit{map-agreement} sub-algorithm (\Cref{map-algo}); the map-agreement sub-algorithm may subsequently commit a mapping $c\mapsto \mathcal{D}'$ for $c$, in which case $\mathcal{D}'$ becomes the dependency set of $c$;
we say that $c$ is executable when all its transitive dependencies are known. %no command $c'$ for which no mapping has been committed is reachable from $c$ by recursively following committed mappings. 
Finally, when $c$ is executable, the $p$ may execute $c$ by deciding the c-struct obtained by \emph{locally} running a \emph{graph-processing} algorithm described in \Cref{abstract-algo}.
Any other process can run the three phases independently of $p$ and eventually execute the command despite a crash of $p$.
However, we show that a leaderless  generalized consensus algorithm is always safe and is live when its underlying dependency-set and map-agreement algorithms are safe and live.

@@

A dependency-set algorithm exposes the input interface $\opn{announce}_p\left( c \right)$, to announce a command $c$, the output interface $\opn{suggest}_p \left( c,\mathcal{D}\right)$, to suggest a set of dependencies $\mathcal{D}$ for $c$, and an input interface $\opn{commit}\left( c,\mathcal{D} \right)$ (meant to observe the map-agreement sub-algorithm), where $p\in P$ is the process executing the call.
Moreover, a dependency-set algorithm must ensure that:
\begin{itemize}[noitemsep,nolistsep]
    \item[-] Safety: (\setword{S1}{dep-valid}), for any suggestion $\opn{suggest}_p \left( c,\mathcal{D}\right)$, every command in $\{c\}\cup \mathcal{D}$ must have been announced before; (\setword{S2}{suggestion-agreement}), for any two suggestions $\opn{suggest}_p \left( c_1,\mathcal{D}_1\right)$ and $\opn{suggest}_q \left( c_2,\mathcal{D}_2\right)$, if $c_1$ and $c_2$ do not commute then either $c_1\in \mathcal{D}_2$ or $c_2 \in \mathcal{D}_1$.
%    \item If $c$ was announced by $p$, then eventually a dependency set is suggested for $c$ to $p$.
    \item[-] Liveness: (\setword{L1}{eventual-suggestion}) If a suggestion $\opn{suggest}_p\left( c \right)$ was made, then eventually a suggestion $\opn{suggest}\left( c,\mathcal{D} \right)_p$ is made, and (\setword{L2}{re-suggest}), if a suggestion $\opn{suggest}_p(c,\mathcal{D})$ has been made and no input $\opn{commit}_q\left( c,\mathcal{D'} \right)$ is observed, then eventually a new suggestion $\opn{suggest}_r\left( c,\mathcal{D''} \right)$ for $r\neq p$ is made.
\end{itemize}

An example of implementation of such an algorithm is depicted by the pseudocode of Figures~\ref{fig:deps-set-algo1} and~\ref{fig:deps-set-algo2}. Due to space constraints, our example focuses on satisfying the safety property only. The main idea is the following: a process $p$ announcing a command $c$ broadcasts $c$ to all the processes, and it suggests the dependency-set $deps_p[c]$ for $c$ as the union of the sets $\mathcal{D}$ that it receives as replies from a quorum of $\mathcal{Q}$ processes; on the other hand, a process that receives an announcement for $c$, computes its reply $\mathcal{D}$ for $c$ as the set of commands $d$ that it received so far, and such that $c$ and $d$ do not commute. The choice of the value of $\mathcal{Q}$ must be such that the intersection of any two quorums is not empty. About correctness, we can easily prove that such an implementation satisfies the safety property by noticing that:  any process $p$ inserts a command $d$ in a dependency-set $\mathcal{D}$ that $p$ sent as a reply for a command $c$, only if $p$ previously received $d$. Therefore, every command suggested in a dependency-set must have been announced before. Further, since a process can suggest a set of dependencies $deps_p[c]$ for a command $c$ after it has received a reply for $c$ from at least a quorum of processes, and thanks to the requirement that we enforced on the intersection of quorums, we can conclude that for any two suggestions $\opn{suggest}_p \left( c_1,deps_p[c_1]\right)$ and $\opn{suggest}_q \left( c_2,deps_q[c_2]\right)$,  if $c_1$ and $c_2$ do not commute then either $c_1\in deps_q[c_2]$ or $c_2 \in deps_p[c_1]$.

@@

		\begin{algorithmic}[1]
			{\scriptsize
			\START[$\opn{init}_p$]{{\bf upon}}{$()$}
			\State $m_p\gets \emptyset$
			\State $\sigma_p\gets \bot$
			\END
			\START[$\opn{GC-propose}_p$]{{\bf upon}}{$(c)$}
			\State {\bf call} $\opn{announce}_p\left( c \right)$
			\END
			\Statex
			\START[$\opn{suggest}_p$]{{\bf upon}}{$(c,\mathcal{D})$}
			\State  $\opn{propose}_p( c,\mathcal{D})$ 			
			\END
			\Statex
			\START[$\opn{commit}_p$]{{\bf upon}}{$(c,\mathcal{D})$}
			\State  $m_p \gets m_p \oplus \left[ c \mapsto \mathcal{D}\right]$	
			\State $\sigma_p \gets  \opn{c-struct}\left(m_p\right)$
			\State $\opn{GC-decide}_p(\sigma_p)$		
			\END
			}
		\end{algorithmic}
	\end{varwidth}%
}
\label{fig:deps-set-algo3}
}
\hspace*{0.1in}\subfigure[Example of dependency-set algorithm implementation - part 1.]{
\noindent\fbox{%
	\begin{varwidth}{\dimexpr\linewidth-2\fboxsep-2\fboxrule\relax}

		\begin{algorithmic}[1]
			{\scriptsize
			\START[\texttt{init$_p$}]{{\bf upon}}{$()$} \label{alg:init}
			\State $cmds_p\gets\emptyset$ \label{alg:init-cmds}
			\State $\forall c,$ $deps_p[c]\gets \emptyset$ \label{alg:init-deps}
			\State $\forall c,$ $heard_p[c]\gets \emptyset$ \label{alg:init-heard}
			\State $\forall c,$ $suggested_p[c]\gets false$ \label{alg:init-suggested}
			\END
			\Statex
			\START[\texttt{announce$_p$}]{{\bf upon}}{$(c)$} \label{alg:announce}
			\State $cmds_p\gets cmds_p\cup\{c\}$ \label{alg:add-command}
			\State $\forall q\in P$ {\bf send} $\langle \texttt{cmd}$ $|$ $c \rangle$ {\bf to} $q$\label{alg:send-new-cmd}
			\END
			}
		\end{algorithmic}
	\end{varwidth}%
}
\label{fig:deps-set-algo1}
}	
\hspace*{0.1in}\subfigure[Example of dependency-set algorithm implementation - part 2.]{
\noindent\fbox{%
	\begin{varwidth}{\dimexpr\linewidth-2\fboxsep-2\fboxrule\relax}

		\begin{algorithmic}[1]
			\makeatletter
			\setcounter{ALG@line}{8}
			\makeatother
			{\scriptsize
			\START[$\langle \texttt{cmd}$ $|$ $c \rangle$]{{\bf upon receive}}{ {\bf from} $q$} \label{alg:receive-new-cmd}
			\State $\mathcal{D} \gets \{ d : d \in cmds_p \land$\label{alg:update-d}
			\Statex \hspace{2cm}$\land d\neq c \land d \asymp c \}$
			\State $cmds_p\gets cmds_p\cup\{c\}$\label{alg:update-cmds-p}
			\State {\bf send} $\langle \texttt{deps}$ $|$ $c,\mathcal{D} \rangle$ {\bf to} $q$\label{alg:send-deps}
			\END
			\Statex
			\START[$\langle \texttt{deps}$ $|$ $c,\mathcal{D} \rangle$]{{\bf upon receive}}{ {\bf from} $q$} \label{alg:receive-deps}
			\State $deps_p[c]\gets deps_p[c]\cup \mathcal{D}$\label{alg:update-deps-p}
			\State $heard_p[c]\gets heard_p[c]\cup \{q\}$
			\END
			\Statex
			\START[]{{\bf upon}}{$\exists c$ $:$ $ | heard_p[c] | \geq \mathcal{Q}$ $\land$} \label{alg:receive-deps}
			\Statex \hspace{1cm}$\land$ $suggested_p[c]=false$
			\State $suggested_p[c]\gets true$
			\State {\bf call} $\texttt{suggest}_p(c,deps_p[c])$ \label{alg:call-suggest}
			\END
			}
		\end{algorithmic}
	\end{varwidth}%
}
\label{fig:deps-set-algo2}
}
\caption{Leaderless generalized consensus, and example of dependency-set algorithm implementation.}
\label{fig:deps-set-algo}
\end{figure*}
